Relational Parametricity for Algebraically Indexed Types

### 2012-06-05 paper/section-general-framework.tex
Minor changes to the discussion of the syntax of indexing theories

The indexing terms and types of our system are derived from the
following data:
\begin{enumerate}
\item A collection $\SortSet$ of index sorts. We will use the
  meta-syntactic variables $s,s_1,s_2,...$ and so on to stand for
  arbitrary sorts taken from $\SortSet$.
\item A collection $\IndexOpSet$ of index operations, with an
  accompanying function $\indexOpArity : \IndexOpSet \to \SortSet^*
  \times \SortSet$ (we use the notation $A^*$ to denote the set of
  lists of elements of some set $A$). 
\item A collection $\PrimTypeSet$ of primitive types, with an
  accompanying function $\primTyArity : \PrimTypeSet \to
  \SortSet^*$.
\end{enumerate}

@@

We also assume a countably infinite collection of index expression
variables. We will use meta-syntactic variables $i, i_1, i_2, ...$ and
so on to stand for index expression variables. \emph{Index contexts}
$\Delta = i_1 : s_1, ..., i_n : s_n$ are built from the index
expression variable/sort pairs, such that all the index variable names
are distinct.

@@

  $\Delta = i_1 : s_1, ..., i_n : s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.

@@

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index terms $\Delta \vdash e : s$
and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are only well-indexed with
respect to some index context $\Delta$. The rules for well-sorted
index terms are particularly simple: either an index term is a
variable that appears in the context (rule \TirName{IVar}), or it is
an application of an index operation taken from $\IndexOpSet$ to other
index terms (rule \TirName{IOp}). The rules for well-indexed types
include the usual rules for constructing types of the simply-typed
$\lambda$-calculus with unit, sum and tuple types (rules
\TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index terms
$e_1,...,e_n$, the well-indexed type $\tyPrim{X}{e_1,...,e_n}$. The
rule \TirName{TyForall} permits the formation of universally
quantified types, where the universal quantification ranges over all
index terms of some sort.

@@

It will technically convenient to express substitution of index terms
in our framework in terms of simultaneous substitutions.  Given a pair
of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1, ..., i_n
\mathord: s_n$, a \emph{simultaneous substitution} $\sigma : \Delta
\Rightarrow \Delta'$ is a sequence of terms $\sigma = (e_1,...,e_n)$
such that for each $j$ such that $1 \leq j \leq n$, $\Delta \vdash e_j
: s_j$. Given a simultaneous substitution $\sigma = (e_1,...,e_n) :
\Delta \Rightarrow \Delta'$ and a variable $i_j \mathord: s_j$ in
$\Delta'$, we write $\sigma(i_j)$ for the index term $e_j$.

For a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/type pair $i\mathord:s$ such that $i$ does not appear in
$\Delta$, we define the \emph{projection} simultaneous substitution
$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
$\pi_{i\mathord:s} = (i_1,...,i_n)$. Note that the subscript on
$\pi_{i\mathord:s}$ is the variable/type pair that is being projected
away.

For a simultaneous substitution $\sigma : \Delta \Rightarrow \Delta'$
where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
variable/type pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
simultaneous substitution $\sigma_{i\mathord:s} = (\sigma(i_1), ...,
\sigma(i_n), i) : \Delta,i\mathord:s \Rightarrow
\Delta',i\mathord:s$. Note that we have implicity used the fact that
well-sortedness of index terms is preserved by weakening.

Given a simultaneous substitution $\sigma : \Delta \Rightarrow
\Delta'$ and a well-sorted index term $\Delta' \vdash e : s$, we can
apply $\sigma$ to $e$ to produce a new index term $\sigma^*e$. The new
index term $\sigma^*e$ is defined by the following two clauses:
\begin{mathpar}
  \sigma^*i \isDefinedAs \sigma(i)

  \sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs \indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)
\end{mathpar}
Similarly, given a well-indexed type $\Delta' \vdash A \isType$, we
can apply $\sigma$ to $A$ to produce a new type $\sigma^*A$. The
application of a simultaneous substitution to a type is defined
recursively by the following clauses:
\begin{mathpar}
  \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}

@@

The \emph{composition} of two simultaneous substitutions $\sigma :
\Delta \Rightarrow \Delta'$ and $\sigma' : \Delta' \Rightarrow
\Delta''$, where $\sigma' = (e'_1,...,e'_n)$, is defined as $\sigma'
\circ \sigma \isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) : \Delta
\Rightarrow \Delta''$.

>>>>

The indexing terms and types of an instantiation of our general
framework are derived from the following data:
\begin{enumerate}
\item A collection $\SortSet$ of index sorts. We will use the
  meta-syntactic variables $s,s_1,s_2,...$ and so on to stand for
  arbitrary sorts taken from $\SortSet$.
\item A collection $\IndexOpSet$ of index operations, with a function
  $\indexOpArity : \IndexOpSet \to \SortSet^* \times \SortSet$ (we use
  the notation $A^*$ to denote the set of lists of elements of some
  set $A$).
\item A collection $\PrimTypeSet$ of primitive types, with a function
  $\primTyArity : \PrimTypeSet \to \SortSet^*$, describing the sorts
  of the arguments of each primitive type.
\end{enumerate}

\begin{example}[Origin Invariance]
  \fixme{Do this, then update the units of measure example}
\end{example}

@@

We assume a countably infinite collection of index variable names $i,
i_1, i_2,$ and so on. \emph{Index contexts} $\Delta = i_1 \mathord:
s_1, ..., i_n \mathord: s_n$ are lists of index expression
variable/sort pairs, such that all the index variable names are
distinct.

@@

  $\Delta = i_1 \mathord: s_1, ..., i_n \mathord: s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.

@@

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index terms $\Delta \vdash e : s$
and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index terms are particularly simple: either an index term is a
variable that appears in the context (rule \TirName{IVar}), or it is
an application of an index operation taken from $\IndexOpSet$ to other
index terms (rule \TirName{IOp}). The rules for well-indexed types
include the usual rules for constructing types of the simply-typed
$\lambda$-calculus with unit, sum and tuple types (rules
\TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index terms
$e_1,...,e_n$, the well-indexed type $\tyPrim{X}{e_1,...,e_n}$. The
rule \TirName{TyForall} permits the formation of universally
quantified types, where the universal quantification ranges over all
index terms of some sort.

@@

It will be technically convenient to express substitution of index
terms in our framework in terms of simultaneous substitutions.  Given
a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1,
..., i_n \mathord: s_n$, a \emph{simultaneous substitution} $\Delta
\vdash \sigma \Rightarrow \Delta'$ is a sequence of terms $\sigma =
(e_1,...,e_n)$ such that for each $j$ such that $1 \leq j \leq n$,
$\Delta \vdash e_j : s_j$. Given a simultaneous substitution $\Delta
\vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a variable $i_j
\mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for the index term
$e_j$.

Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/type pair $i\mathord:s$ such that $i$ does not appear in
$\Delta$, we define the \emph{projection} simultaneous substitution
$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
$\pi_{i\mathord:s} = (i_1,...,i_n)$. The subscript on
$\pi_{i\mathord:s}$ is the variable/type pair that is being discarded.

For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$ where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
variable/type pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
simultaneous substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. Note that we have implicity used the fact that
well-sortedness of index terms is preserved by addition of extra items
to the context.

Application of a simultaneous substitution $\Delta \vdash \sigma
\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
: s$. The expression $\sigma^*e$ is defined by the following two
clauses:
\begin{mathpar}
  \sigma^*i \isDefinedAs \sigma(i)

  \sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs \indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)
\end{mathpar}
Similarly, given a well-indexed type $\Delta' \vdash A \isType$, we
can apply $\sigma$ to $A$ to produce a new type $\sigma^*A$. The
application of a simultaneous substitution to a type is defined
by the following clauses:
\begin{mathpar}
  \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}

@@

The \emph{composition} of two simultaneous substitutions $\Delta
\vdash \sigma \Rightarrow \Delta'$ and $\Delta' \vdash \sigma'
\Rightarrow \Delta''$, where $\sigma' = (e'_1,...,e'_n)$, is defined
as $\Delta \vdash \sigma' \circ \sigma \isDefinedAs (\sigma^*e'_1,
..., \sigma^*e'_n) \Rightarrow \Delta''$.
